---
title: "Cradle Analysis, MSc Diss"
author: "Amy Hatton, UCL"
date: "26/07/2020"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

Setup and install libraries

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir=rprojroot::find_rstudio_root_file())

library(spatstat)
library(raster)
library(maptools)
library(rgdal)
library(sf)
library(viridisLite)
library(tmap)
library(tidyverse)
library(tmaptools)
library(here)
library(sp)

#set up the south african projection system
rsa_grid <- CRS(
  "+proj=tmerc +lat_0=0 +lon_0=27 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
wgs84_latlong <- st_crs(4326)

```

## Point pattern analysis with spatstat:

Create a multitype pattern

```{r}
sites_analysis <- st_read("data/cradle/sites_analysis.gpkg")
sites_analysis$site_category <- as.factor(sites_analysis$site_category)

#set intersection of malmani dolomite and cradle as owin
coh_malmani <- st_transform(st_read("data/cradle/obbes_malmani.gpkg"), rsa_grid)
#set malmani_coh as window for analysis
malmani_owin <- as.owin(coh_malmani)

#convert to sp and the to ppp object
caves_sp <- as(sites_analysis, "Spatial")
library(spbabel)
coords_temp <- spbabel::sptable(caves_sp)
caves_ppp <- ppp(x = coords_temp$x_, y= coords_temp$y_,
                 window = malmani_owin, marks = caves_sp$site_category)
```

Split according to site category and explore the spatial distribution:

```{r}
all_caves <- as.ppp(coordinates(
  caves_sp[caves_sp$site_category=="all_cave", ]),
  as.owin(malmani_owin))

macro_caves <- as.ppp(coordinates(
  caves_sp[caves_sp$site_category=="macro_fossil_cave", ]),
  as.owin(malmani_owin))

hom_caves <- as.ppp(coordinates(
  caves_sp[caves_sp$site_category=="hominin_cave", ]),
  as.owin(malmani_owin))

```

### Kernel Density Estimation

First calculate the mean nearest neighbour distance, to determine the
kernel bandwidth:

all cave mean nnd= 522m macro cave mean nnd= 581m hominin cave mean nnd=
756m

```{r}
nndc <- nndist(caves_ppp, by=marks(caves_ppp))
aggregate(nndc, by=list(from=marks(caves_ppp)), mean)
nnd_list <- c(756, 581, 522)
nnd_list*3
```

Density of sites in the cradle of humankind two main clusters , one in
the north, one in the south The sites also seem to be aligned along an
axis that is more north-south than the dolomite

```{r}
# Overall density of all points in our data
d.hom.1 <- density.ppp(hom_caves, sigma = 500)
d.hom.2 <- density.ppp(hom_caves, sigma = 2268)
d.hom.3 <- density.ppp(hom_caves, sigma = 6000)
d.macro.1 <- density.ppp(macro_caves, sigma = 500)
d.macro.2 <- density.ppp(macro_caves, sigma = 1743)
d.macro.3 <- density.ppp(macro_caves, sigma = 6000)
d.all_caves.1 <- density.ppp(all_caves, sigma = 500)
d.all_caves.2 <- density.ppp(all_caves, sigma = 1566)
d.all_caves.3 <- density.ppp(all_caves, sigma = 6000)
```

\#\#\#Plot kernel density estimates

Density of all cave sites is mostly in northern edge, while density of
hominin cave sites is in southern edge

```{r, fig.width= '33%'}

png("figs/coh_chp/KDE.png", width = 1300, height = 1200, res =150)
par(mfrow = c(3,3), mar = c(0.7, 0.4, 1, 1.2))#bottom, left, top, and right
#threshold = 1000m
plot(d.all_caves.1, main = "All cave sites") # Plot interpolated values
plot(as.ppp(all_caves), add = TRUE, cex = 0.9, pch =1)

plot(d.macro.1, main = "Macro fossil cave sites") 
plot(as.ppp(macro_caves), add = TRUE, cex = 1, pch =1)

plot(d.hom.1, main = "Hominin cave sites") # Plot interpolated values
plot(hom_caves, add = TRUE, cex = 1, pch =1)
par(mar = c(0.4, 0.4, 0.4, 1.2))
#threshold = 3000m
plot(d.all_caves.2, main = "") 
plot(as.ppp(all_caves), add = TRUE, cex = 1, pch =1)

plot(d.macro.2, main = "") 
plot(as.ppp(macro_caves), add = TRUE, cex = 1, pch =1)

plot(d.hom.2, main = "") # Plot interpolated values
plot(hom_caves, add = TRUE, cex = 0.8, pch =1)

#threshold = 5000m
plot(d.all_caves.3, main = "") # Plot interpolated values
plot(as.ppp(all_caves), add = TRUE, cex = 1, pch =1)

plot(d.macro.3, main = "") # Plot interpolated values
plot(as.ppp(macro_caves), add = TRUE, cex = 1, pch =1)

plot(d.hom.3, main = "") # Plot interpolated values
plot(hom_caves, add = TRUE, cex = 1, pch =1)

dev.off()
```

#### investigating anisotropy (directional layout of point - geological fractures)

Visualise:

Fry plots:

```{r, fig.show="hold", fig.width='33%'}
png("figs/coh_chp/fry_plot.png", width = 1300,
    height = 400, res =150)
par(mfrow = c(1,3), mar = c(0.7, 0.4, 0.7, 0.7))

fryplot(all_caves, cex=0.3, axes=T, main="All caves")
fryplot(macro_caves, cex=0.3, axes=T, main="Macro caves")
fryplot(hom_caves, cex=0.3, axes=T, main="Hominin caves")
dev.off()
```

Rose Diagrams

calculate a pairorient() which is the probability distribution of angles

```{r, fig.show='hold', fig.width='33%'}
png("figs/coh_chp/rose_plot.png", width = 1300,
    height = 400, res =150)
par(mfrow = c(1,3), mar = c(0.7, 0.4, 0.7, 0.7))

rose_ac <- pairorient(all_caves,r1=500, r2=3000, sigma=4)
rose(rose_ac,  start=270, clockwise=TRUE, col = "gray", main= "All caves")
rose_mc <- pairorient(macro_caves,r1=500, r2=3000, sigma=4)
rose(rose_mc,  start=270, clockwise=TRUE, col = "gray", main= "Macro fossil caves")
rose_hc <- pairorient(hom_caves,r1=500, r2=3000, sigma=4)
rose(rose_hc,  start=270, clockwise=TRUE, col = "gray", main= "Hominin fossil caves")

dev.off()
```

## Investigate the intensity of point for each of the covariates

### Covariate data

```{r}
grids <- list.files(path="data/cradle/covariates/", pattern = "*.tif$")
rs <- raster::stack(paste0("data/cradle/covariates/", grids))
library(tools)
names <- file_path_sans_ext(list.files("data/cradle/covariates/"))
names(rs) <- c(names)
```

Convert to im objects:

```{r}
dykes.im <- as.im(as(rs$dist_dykes,"SpatialGridDataFrame"))
faults.im <- as.im(as(rs$dist_faults,"SpatialGridDataFrame"))
lith.im <-  as.im(as(rs$dist_lith_bound,"SpatialGridDataFrame"))
streams.im <- as.im(as(rs$dist_streams, "SpatialGridDataFrame"))
dem.im <- as.im(as(rs$malmani_dem, "SpatialGridDataFrame"))
tpi.im <- as.im(as(rs$tpi, "SpatialGridDataFrame"))  
tri.im <- as.im(as(rs$tri, "SpatialGridDataFrame"))
```

create rhohat plots

```{r}
rh_dykes <- rhohat(all_caves, dykes.im, confidence = 0.95)
rh_dykes_hom <- rhohat(hom_caves, dykes.im, confidence = 0.95)
rh_faults <- rhohat(all_caves, faults.im, confidence = 0.95)
rh_fauts_hom <- rhohat(hom_caves, faults.im, confidence = 0.95)
rh_lith <- rhohat(all_caves, lith.im, confidence = 0.95)
rh_lith_hom <- rhohat(hom_caves, lith.im, confidence = 0.95)
rh_streams_hom <- rhohat(hom_caves,
                           streams.im, confidence = 0.95)
rh_streams <- rhohat(all_caves,
                       streams.im, confidence = 0.95)

rh_dem <- rhohat(all_caves, dem.im, confidence = 0.95)
rh_dem_hom <- rhohat(hom_caves, dem.im, confidence = 0.95)

rh_tpi<- rhohat(all_caves, tpi.im, confidence = 0.95)
rh_tpi_hom <- rhohat(hom_caves, tpi.im, confidence = 0.95)

rh_tri <- rhohat(all_caves, tri.im, confidence = 0.95)
rh_tri_hom <- rhohat(hom_caves, tri.im, confidence = 0.95)
```

Density plot for all caves for the covariates: *Elevation:* the
elevation in the study area is restricted to between 1350 and 1550 masl,
there are two peaks for the intensity of sites at about 1430m and 1510m
(does this correspond to the info from Marker and Moon 1969?)

*Heat Load Index:* The intensity of cave sites is fairly stable for all
ranges of the HLI (0-1), although there is a small peak at about 0.4

*Terrain Ruggedness Index:* The TRI within the study area is very
constrained and represent a very level terrain surface. Values between
1-80 are considered a level terrain surface and the values in the study
area are between 0-70, with cave sites located in the 0-40 range. The
intensity of sites over this range is consistent - not a good predictor
of site location

*Distance From Streams (calculated with a threshold of 500):* The
intensity of sites is consistent for all distances to stream at this
threshold. - not a good predictor

*Distance From Streams (calculated with a threshold of 50000):* Site
intensity s highest at closer distance to streams calculated at this
threshold, with intensity dropping after about 2000m distance from
streams.

Figure out what ranges to use for the plots (range of values across study area)

```{r}
#1200 to 1700m
hist.im(elev.im)
#0-70
hist.im(tri.im)
# 0 -19000
hist.im(streams.im)
#0 - 12000
hist.im(dykes.im)
```

#### Plots of the covarites to place next to each rhohat plot

Create Hillshade as an underlay

```{r}
slope <- terrain(rs$malmani_dem, opt="slope", unit="radians",
                 neighbors = 8)
aspect <- terrain(rs$malmani_dem, opt="aspect", unit="radians",
                 neighbors = 8)
hs <- hillShade(slope, aspect, angle = 45, direction = 315)
```


#### base r rhohat plot maps

```{r}
library(RColorBrewer)
dpi=300

#elevation
pdf("figs/rhohat_plots.pdf", width =8, height = 14)
par(mfrow=c(4,3))
par(mar=c(4.1,3.1,3.1,2.1))
plot(rh_dem, main = "All Caves", xlab="Elevation (m)",xlim=c(1200,1700), legend=FALSE)
plot(rh_dem_hom, main = "Hominin Caves", xlab="Elevation (m)",xlim=c(1200,1700), legend=FALSE)
par(mar=c(2.1,2.1,2.1,2.1))
plot(rs$malmani_dem, main="Elevation", col=inferno(10), axes=FALSE, box=FALSE, legend.args=list(text="",side=2), horizontal=TRUE)

#TPI
par(mar=c(4.1,3.11,3.1,2.1))
plot(rh_tpi, main = "All Caves", xlab="Topographic Position Index",xlim=c(-6, 6), legend=FALSE)
plot(rh_tpi_hom, main = "Hominin Caves", xlab="Topographic Position Index",xlim=c(-6,6), legend=FALSE)

#make a diverging colour pallette for TPI
cols <- brewer.pal(3, "PRGn")
pal <- colorRampPalette(cols)
par(mar=c(2.1,2.1,2.1,2.1))
plot(rs$tpi, main="TPI", col=pal(12), axes=FALSE, box=FALSE, legend.args=list(text="",side=2), horizontal=TRUE)

#TRI
par(mar=c(4.1,3.11,3.1,2.1))
plot(rh_tri, main = "All Caves", xlab="Terrain Ruggedness Index", xlim=c(0,15), legend=FALSE)
plot(rh_tri_hom, main = "Hominin Caves", xlab="Terrain Ruggedness Index", xlim=c(0,15), legend=FALSE)
par(mar=c(2.1,2.1,2.1,2.1))
plot(rs$tri, main="TRI", col=, axes=FALSE, box=FALSE, legend.args=list(text="",side=2), horizontal=TRUE)

#streams
par(mar=c(4.1,3.11,3.1,2.1))
plot(rh_streams, main = "All Caves",
     xlab="Distance to Streams (m)", xlim=c(0,13000), legend=FALSE)
plot(rh_streams_hom, main = "Hominin Caves",
     xlab="Distance to Streams (m)", xlim=c(0,13000), legend=FALSE)
par(mar=c(2.1,2.1,2.1,2.1))
plot(rs$dist_streams, main="Distance to Streams", col=inferno(10), axes=FALSE, box=FALSE, legend.args=list(text="",side=2), horizontal=TRUE)

dev.off()
```

```{r}
library(RColorBrewer)
dpi=300

#Lithological boundaries
pdf("figs/rhohat_plots1.pdf", width =8, height = 12)
par(mfrow=c(3,3))
par(mar=c(4.1,3.1,3.1,2.1))
plot(rh_lith, main = "All Caves", xlab="Distance to Lithological boundaries (m)",xlim=c(0,3000), legend=FALSE)
plot(rh_lith_hom, main = "Hominin Caves", xlab="Distance to Lithological boundaries (m)",xlim=c(0,3000), legend=FALSE)
par(mar=c(2.1,2.1,2.1,2.1))
plot(rs$dist_lith_bound, main="Distance to\nLithological boundaries", col=inferno(10), axes=FALSE, box=FALSE, legend.args=list(text="",side=2), horizontal=TRUE)

#dykes
par(mar=c(4.1,3.11,3.1,2.1))
plot(rh_dykes, main = "All Caves", xlab="Distance to Dykes",xlim=c(0, 6000), legend=FALSE)
plot(rh_dykes_hom, main = "Hominin Caves", xlab="Distance to Dykes",xlim=c(0,6000), legend=FALSE)
plot(rs$dist_dykes, main="Distance to Dykes", col=inferno(10), axes=FALSE, box=FALSE, legend.args=list(text="",side=2), horizontal=TRUE)

#faults
par(mar=c(4.1,3.11,3.1,2.1))
plot(rh_faults, main = "All Caves", xlab="Distance to faults", xlim=c(0,8000), legend=FALSE)
plot(rh_faults_hom, main = "Hominin Caves", xlab="Distance to faults", xlim=c(0,8000), legend=FALSE)
par(mar=c(2.1,2.1,2.1,2.1))
plot(rs$dist_faults, main="Distance to Faults", col=, axes=FALSE, box=FALSE, legend.args=list(text="",side=2), horizontal=TRUE)

dev.off()
```


## Analysis: Model fitting

Check for collinearity among the variables

## There is no intercorrelation among variables at 0.75 cutoff

```{r}
library(virtualspecies)
removeCollinearity(rs, multicollinearity.cutoff = 0.75,
                   select.variables = FALSE, sample.points = TRUE,
                   nb.points = 100000,
                   plot = TRUE)
```

```{r}
#convert to im and stack, remove non-stacked variables
vars <- list(as.im(as(rs[[1]],"SpatialGridDataFrame")),
             as.im(as(rs[[2]],"SpatialGridDataFrame")),
             as.im(as(rs[[3]],"SpatialGridDataFrame")),
             as.im(as(rs[[4]],"SpatialGridDataFrame")),
             as.im(as(rs[[5]], "SpatialGridDataFrame")),
             as.im(as(rs[[6]], "SpatialGridDataFrame")),
             as.im(as(rs[[7]], "SpatialGridDataFrame"))
             )
             
names(vars) <- c("elev","tpi", "tri", "dist_50000", "dykes", "faults", "lith_bound")

```

### Create a model to describe each of the patterns based on covariates

####First order effects

```{r}
library(spatstat)
c_trend <- ~ elev+tpi+ tri+ dist_50000+ dykes+ faults+ lith_bound # Define spatial\

fitc <- step(ppm(all_caves, trend=c_trend, covariates=vars, correction="iso", method="logi"), k=log(5))
summary(fitc)

```

Univariate model for macro fossil caves: all covariate were kept for
final

```{r}
fitm <- step(ppm(macro_caves, trend = c_trend, covariates=vars, correction = "iso", method="logi"), k=log(5))

summary(fitm)

```

Univariate model for hominin fossil caves: Model depends on external
covariates 'elev', 'hli', 'dist\_500' and 'dist\_50000', Terrain
Ruggednes was not included.

```{r}
fith <- step(ppm(hom_caves, trend = c_trend, covariates=vars, correction = "iso", method="logi"), k=log(5))

summary(fith)
```

First order was able to explain all the variation
didn't fit a first + second order model

## Analysis: Model testing

# Based on 999 simulations of CSR.

# Null

```{r}
null <- envelope(caves_ppp, fun=pcfinhom, nsim = 999, divisor="d", correction = "iso")


# First order all caves

#create simulated patterns based on first order for all caves
c_1 <- envelope(fitc, fun = pcfinhom, nsim = 999, correction="iso", divisor = "d", savepatterns=TRUE)

#create envelope for hominin caves based on all caves first order fit
simfitc <- envelope(hom_caves, fun = pcfinhom, nsim = 999, correction="iso", divisor = "d", 
                    simulate=c_1)

# First order macro caves
m_1 <- envelope(fitm, fun=pcfinhom, nsim = 999,correction="iso", divisor="d", savepatterns=TRUE)

simfitm <- envelope(macro_caves, fun=pcfinhom, nsim = 999,
                    correction="iso", divisor="d", simulate=m_1)

# First order hominin caves

simfith <- envelope(fith, fun = pcfinhom, nsim = 999, correction="iso", divisor = "d")
```

## Model plots

### Predictions maps based on models

Create raster based on model fit

```{r}
logodds_caves <- raster(-15.05964000  +(vars[[1]]*0.00098243)+(vars[[2]]*0.07562973)+(vars[[3]]*0.20667870)+(vars[[4]]*-0.00022537)+(vars[[5]]*0.00007790)+(vars[[6]]*-0.00017696 +(vars[[7]]*-0.00002233)), 
                      crs=rsa_grid)

logodds_mac <- raster(-7.51372290 +(vars[[1]]*-0.00550579)+ (vars[[2]]*0.01377018)+(vars[[3]]*0.31368157)+(vars[[4]]*-0.00027172)+(vars[[5]]*0.00036743)+ (vars[[6]]*-0.00007227)+ (vars[[7]]*0.00012990), crs=rsa_grid)

logodds_hom <- raster( -0.241866676 + (vars[[1]]*-0.010860565)+(vars[[3]]*0.095577452)+(vars[[4]]*-0.000438203)+(vars[[6]]*0.000522995)+(vars[[7]]*0.000285699),crs=rsa_grid)

```

### PCF inhomogenous based on models

```{r}
pdf("figs/coh_chp/fitted_models_with_macro.pdf", width =7, height =11)
par(mfrow=c(3,3))

par(mar=c(5.1,5.1,5.1,2.1))
plot(null, legend=FALSE, main="Null", lwd=1.5)

par(mar=c(5.1,5.1,5.1,2.1))
plot(simfitc, legend=FALSE, main="First-order all caves", lwd=1.5)

par(mar=c(5.1,5.1,5.1,2.1))
plot(simfitm, legend=FALSE, main="First-order macro caves", lwd=1.5)

par(mar=c(5.1,5.1,5.1,2.1))
plot(simfith, legend=FALSE, main="First-order hominin caves",lwd=1.5)

par(mar=c(2.1,3.1,2.1,2.1))
plot(logodds_caves, main="Predicted intensity\nall caves", col=inferno(10), axes=FALSE, box=FALSE, legend.args=list(text="",side=2), horizontal=TRUE)

par(mar=c(2.1,3.1,2.1,2.1))
plot(logodds_mac, main="Predicted intensity\nmacro caves", col=inferno(10), axes=FALSE, box=FALSE, legend.args=list(text="",side=2), horizontal=TRUE)

par(mar=c(2.1,3.1,2.1,2.1))
plot(logodds_hom, main="Predicted intensity\nhominin caves", col=inferno(10), axes=FALSE, box=FALSE, legend.args=list(text="",side=2), horizontal=TRUE)

dev.off()
```