---
title: "Griqua Cluster Spatial Analysis, MSc Diss 2020"
author: "Amy Hatton, UCL"
date: "29/09/2020"
output: html_document
---
Set the root of project folder as directory
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir=rprojroot::find_rstudio_root_file())
```
Load libraries
```{r, include=FALSE}
library(spatstat)
library(raster)
library(maptools)
library(rgdal)
library(sf)
library(viridisLite)
library(tmap)
library(tidyverse)
library(tmaptools)
library(here)
library(sp)
#hartebeestheok peojection
lo23 <- CRS(
  "+proj=tmerc +lat_0=0 +lon_0=23 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
#Lat long projection
wgs84_latlong <- st_crs(4326)
```

## Point pattern analysis with spatstat:
Get data and
Create a multitype pattern

```{r}
sites <- st_transform(st_read("data/griqua/kim_sites.gpkg"), lo23)
sites$site_type <- as.factor(sites$site_type)
sites_agecat <- st_transform(st_read("data/griqua/sites_agecat.gpkg"), crs=lo23)
basin <- st_transform(st_read("data/griqua/basins_final.gpkg", layer="boundary"), lo23)
sites_agecat$age_cat <- as.factor(sites_agecat$age_cat)
kim_sites_ages <- st_intersection(sites_agecat, basin)
kim_sites_ages$site_type <- as.factor(kim_sites_ages$site_type)
basin_sp <- as(basin, "Spatial")
#split age cats
young1 <- kim_sites_ages %>% 
  filter(age_cat == "agecat_1"|age_cat == "agecat_2")
mid1 <- kim_sites_ages %>% 
  filter(age_cat == "agecat_3"|age_cat == "agecat_4")
old1 <- kim_sites_ages %>% 
  filter(age_cat == "agecat_5")
#set basin boundary as window for analysis
basin_owin <- as.owin(basin_sp)
#convert to sp and the to ppp object
sites_sp <- as(sites, "Spatial")
sahris_sites_sp <- as(sahris_sites, "Spatial")
young <- as(young1, "Spatial")
mid <- as(mid1, "Spatial")
old <- as(old1, "Spatial")
library(spbabel)
y.coords <- coordinates(young)
m.coords <- coordinates(mid)
o.coords <- coordinates(old)
y_ppp <- ppp(x=y.coords[,1], y=y.coords[,2], window = basin_owin)
m_ppp <- ppp(x=m.coords[,1], y=m.coords[,2], window = basin_owin)
o_ppp<- ppp(x=o.coords[,1], y=o.coords[,2], window = basin_owin)
#with marks
y.ppp <- ppp(x=y.coords[,1], y=y.coords[,2], window = basin_owin, marks=young$site_type)
m.ppp <- ppp(x=m.coords[,1], y=m.coords[,2], window = basin_owin, marks=mid$site_type)
o.ppp<- ppp(x=o.coords[,1], y=o.coords[,2], window = basin_owin, marks=old$site_type)
coords_temp <- spbabel::sptable(sites_sp)
caves_ppp <- as.ppp(coordinates(sites_sp[sites_sp$site_type=="cave"|sites_sp$site_type=="rockshelter",]),
                    as.owin(basin_owin))
open_ppp <- as.ppp(coordinates(sites_sp[sites_sp$site_type=="open-air",]),
                   as.owin(basin_owin))
all_ppp <- as.ppp(coordinates(sites_sp), as.owin(basin_owin))
```

### Kernel Density Estimation

First calculate the mean nearest neighbour distance multiplied by 3, to determine the kernel bandwidth:

Attempt two 

-   old (800-2800ka)= 63114m   
-   middle (130-800ka)= 6603m
-   young (10-130)= 8162m


```{r}
#for (i in list){nndc <- nndist(
 # i, by=marks(i))
#print(mean(nndc))}
nndc <- nndist(caves_ppp, by=marks(caves_ppp))
mean(nndc)
nndc <- nndist(open_ppp, by=marks(open_ppp))
mean(nndc)
nndc <- nndist(sahris_ppp, by=marks(sahris_ppp))
mean(nndc)
nndc <- nndist(y_ppp, by=marks(y_ppp))
mean(nndc)
nndc <- nndist(m_ppp, by=marks(m_ppp))
mean(nndc)
nndc <- nndist(o_ppp, by=marks(o_ppp))
mean(nndc)
#attempt two
nndc <- nndist(o.ppp)
mean(nndc)
nnd_list <- c(3322, 3533,3590,63114, 6603, 8162)
nnd_list*3
```
Empirical distances:
9966  10599  10770 189342  19809  24486
Need to figure out how to do this in a better way (loop through list)

```{r}
# Overall density of all points in our data
#list <- c(caves_ppp, open_ppp, sahris_ppp, a1_2_ppp, a3_4_ppp, a5_ppp)
#ppp_list <- as.solist(list, demote=TRUE)
#for(i in list) {density.ppp(list[i], sigma=3000)}
d.cave.1 <- density.ppp(caves_ppp, sigma = 3000)
d.cave.2 <- density.ppp(caves_ppp, sigma = 9966)
d.cave.3 <- density.ppp(caves_ppp, sigma = 40000)
d.open.1 <- density.ppp(open_ppp, sigma = 3000)
d.open.2 <- density.ppp(open_ppp, sigma = 10599)
d.open.3 <- density.ppp(open_ppp, sigma = 40000)
  
d.old1 <- density.ppp(o_ppp, sigma = 3000)
d.old2 <- density.ppp(o_ppp, sigma = 20000)
d.old3 <- density.ppp(o_ppp, sigma = 40000)
d.mid1 <- density.ppp(m_ppp, sigma = 3000)
d.mid2 <- density.ppp(m_ppp, sigma = 19809)
d.mid3 <- density.ppp(m_ppp, sigma = 40000)
d.young1 <- density.ppp(y_ppp, sigma = 3000)
d.young2 <- density.ppp(y_ppp, sigma = 24486)
d.young3 <- density.ppp(y_ppp, sigma = 40000)
```


#### Plot kernel density estimates

Density of all cave sites is mostly in northern edge, while density of hominin cave sites is in southern edge why??

```{r, fig.width= '33%'}
pdf("figs/griqua/KDE.pdf", width = 7.5, height =12.5)
par(mfrow = c(5,3), mar = c(0.7, 0.4, 0.7, 1.2))#bottom, left, top, and right
#threshold = 1000m
plot(d.cave.1, main = "Cave Sites, small kernel") # Plot interpolated values
plot(as.ppp(caves_ppp), add = TRUE, cex = 1, pch =1)
plot(d.cave.2, main = "Cave Sites, empirical kernel") 
plot(as.ppp(caves_ppp), add = TRUE, cex = 1, pch =1)
plot(d.cave.3, main = "Cave Sites, large kernel") # Plot interpolated values
plot(as.ppp(caves_ppp), add = TRUE, cex = 1, pch =1)
plot(d.open.1, main = "Open-air Sites, small kernel") 
plot(as.ppp(open_ppp), add = TRUE, cex = 1, pch =2)
plot(d.open.2, main = "Open-air Sites, empirical kernel") 
plot(as.ppp(open_ppp), add = TRUE, cex = 1, pch =2)
plot(d.open.3, main = "Open-air Sites, large kernel") # Plot interpolated values
plot(as.ppp(open_ppp), add = TRUE, cex = 1, pch =2)
plot(d.young1, main = "Sites 10-130ka, small kernel") # Plot interpolated values
plot(as.ppp(y.ppp), add = TRUE, cex = 1, pch =c(1,2))
plot(d.young2, main = "Sites 10-130ka, empirical kernel") 
plot(as.ppp(y.ppp), add = TRUE, cex = 1, pch =c(1,2))
plot(d.young3, main = "Sites 10-130ka, large kernel") # Plot interpolated values
plot(as.ppp(y.ppp), add = TRUE, cex = 1, pch =c(1,2))
plot(d.mid1, main = "Sites 130-800ka, small kernel") 
plot(as.ppp(m.ppp), add = TRUE, cex = 1, pch =c(1,2))
plot(d.mid2, main = "Sites 130-800ka, empirical kernel") 
plot(as.ppp(m.ppp), add = TRUE, cex = 1, pch =c(1,2))
plot(d.mid3, main = "Sites 130-800ka, large kernel") # Plot interpolated values
plot(as.ppp(m.ppp), add = TRUE, cex = 1, pch =c(1,2))
plot(d.old1, main = "Sites 800-2800ka, small kernel") # Plot interpolated values
plot(as.ppp(o.ppp), add = TRUE, cex = 1, pch =c(1,2))
plot(d.old2, main = "Sites 800-2800ka, empirical kernel") # Plot interpolated values
plot(as.ppp(o.ppp), add = TRUE, cex = 1, pch =c(1,2))
plot(d.old3, main = "Sites 800-2800ka, large kernel") # Plot interpolated values
plot(as.ppp(o.ppp), add = TRUE, cex = 1, pch =c(1,2))
dev.off()
```

## Distance Based Approaches


### Covariates (environmental variables)

look at the effect of variables on the location of sites

Variables to test :

-    Elevation

-   deviation from north

-   topographic position index

-   terrain ruggedness index

-   distance from streams (large and small threshold)

-   distance from dolomite - geology type ? (or dist from quarternary seds)

-   distance to watershed boundaries -

-   positive openess (viewshed)

```{r}
dem <- raster("data/griqua/kim_dem_exact.tif")

```


## Create im objects

```{r}
library(maptools)
vars <- list(as.im(as(rs[[1]], "SpatialGridDataFrame")),
             as.im(as(rs[[2]], "SpatialGridDataFrame")),
             as.im(as(rs[[3]], "SpatialGridDataFrame")),
             as.im(as(rs[[4]], "SpatialGridDataFrame")),
             as.im(as(rs[[5]], "SpatialGridDataFrame")),
             as.im(as(rs[[6]], "SpatialGridDataFrame")))
names(vars) <- c(names)
```
## Exploring relationship between covariates and intensity (using all sites point pattern)

```{r}
rh_cal <- rhohat(all_ppp, vars[[1]], confidence = 0.95)
rh_dol <- rhohat(all_ppp, vars[[2]], confidence = 0.95)
rh_streams <- rhohat(all_ppp, vars[[3]], confidence = 0.95)
rh_dem <- rhohat(all_ppp, vars[[4]], confidence = 0.95)
rh_ndev <- rhohat(all_ppp, vars[[5]], confidence = 0.95)
rh_tpi <- rhohat(all_ppp, vars[[6]], confidence = 0.95)
rh_tri <- rhohat(all_ppp, vars[[7]], confidence = 0.95)
```


Plot rhohat plots
```{r}
library(RColorBrewer)
pdf("figs/griqua/rhohat_plots_short.pdf", width =15, height = 8)
par(mfrow=c(3,4))
par(mar=c(4.1,4.1,3.1,2.1))
plot(rh_cal, main = "", xlab="Distance to calcrete m)",xlim=c(0,34620), legend=FALSE)
par(mar=c(2.1,2.1,2.1,5.1))
plot(rs[[1]], main="Distance to calcrete (m)", col=inferno(10), axes=FALSE, box=FALSE)
#p3 <- recordPlot()
par(mar=c(4.1,4.1,3.1,2.1))
plot(rh_dol, main = "", xlab="Distance to dolomite (m)",xlim=c(0, 119659), legend=FALSE)
par(mar=c(2.1,2.1,2.1,5.1))
plot(rs[[2]], main="Distance to dolomite (m)", col=inferno(10), axes=FALSE, box=FALSE)
par(mar=c(4.1,4.1,3.1,2.1))
plot(rh_streams, main = "",
     xlab="Distance to streams (m)", xlim=c(0,60977), legend=FALSE)
par(mar=c(2.1,2.1,2.1,5.1))
plot(rs[[3]], main="Distance to streams (m)", col=inferno(10), axes=FALSE, box=FALSE)
par(mar=c(4.1,4.1,3.1,2.1))
plot(rh_dem, main = "",
     xlab="Elevation (m)", xlim=c(998, 1860), legend=FALSE)
par(mar=c(2.1,2.1,2.1,5.1))
plot(rs[[4]], main="Elevation (m)", col=viridis(12), axes=FALSE, box=FALSE)
par(mar=c(4.1,4.1,3.1,2.1))
plot(rh_tpi, main = "", xlab="Terrain Position Index", xlim=c(-40,39), legend=FALSE)
#make a diverging colour pallette for TPI
cols <- brewer.pal(3, "PRGn")
pal <- colorRampPalette(cols)
par(mar=c(2.1,2.1,2.1,5.1))
plot(rs[[6]], main="TPI", col=pal(12), axes=FALSE, box=FALSE)
par(mar=c(4.1,4.1,3.1,2.1))
plot(rh_tri, main = "", xlab="Terrain Ruggedness Index", xlim=c(0,64), legend=FALSE)
par(mar=c(2.1,2.1,2.1,5.1))
plot(rs[[7]], main="TRI", col=, axes=FALSE, box=FALSE)
dev.off()
```

```{r}
library(spatstat)
```

## Model fitting

Check for collinearity

Had to remove slope because it was intercorrelated 

```{r}
library(virtualspecies)
removeCollinearity(rs, multicollinearity.cutoff = 0.75,
                   select.variables = FALSE, sample.points = TRUE, nb.points = 100000,
                   plot = TRUE) # No collinear variables
```
### First order model all sites
```{r}
ktrend <- ~ d_cal + d_dol + d_streams + kim_dem_exact + tpi + tri
fit_all <- step(ppm(all_ppp, trend = ktrend, covariates=vars, correction = "iso", method="logi"), k=log(5))
summary(fit_all)
```
#### First order models for each group of sites


### Model Testing 

## All sites
### Null
```{r}
na <- envelope(all_ppp, fun = pcfinhom, correction="iso",
                    divisor="d", nsim = 999, fix.n=TRUE)
```
### First order
```{r}
fa <- envelope(fit_all, fun = pcfinhom, nsim = 999, correction = "iso", divisor = "d", fix.n=TRUE)
```
### Prediction map
```{r}
logodds_all <- raster(-20.05716
 +(vars[[1]]*-0.00010)+(vars[[2]]*-0.00001)+(vars[[4]]*-0.00081)+(vars[[5]]*-0.04737
)+(vars[[6]]*0.25147), crs=lo23)
```



### Plotting the model testing 

####All sites


### Plot the PCF 
#### Null

#### First order based on pattern of all caves
```{r}
pdf("figs/griqua/fitted_models_first_order_all.pdf", width =2.5, height =12)
par(mfrow = c(5,1))
par(mar=c(2.1,2.1,2.1,2.1))
plot(fa1, legend=FALSE,main="", lwd=1.5, ylim=c(0,15))
plot(fa2, legend=FALSE,main="",  lwd=1.5, ylim=c(0,12))
plot(fa3, legend=FALSE,main="",  lwd=1.5, ylim=c(0,20))
plot(fa4, legend=FALSE,main="" , lwd=1.5, ylim=c(0,15))
plot(fa5, legend=FALSE,main="",  lwd=1.5, ylim=c(0,1))
dev.off()
```